\section{Hardware Implementation and Demonstrations}

\begin{quote}
\textit{Author's Note (Devon): I cannot tell you how satisfying it was to see the Verilog simulation output match the Python VM match the Coq extraction. Three compilation paths---with the Python layer delegating to the extracted OCaml runner rather than reimplementing semantics---written across three different languages, producing the \textbf{same answer}. That's not luck. That's not coincidence. That is the whole isomorphism claim being confirmed in practice. Or at least, the parts of the claim that the current test coverage exercises. That is a distinction I am not allowed to blur.}
\end{quote}

\subsection{Why Hardware Matters}

A computational model is only as credible as its implementation. The Turing Machine was a thought experiment---it was never built as a physical device (though it could be). The Church-Turing thesis claims that any ``mechanical'' computation can be performed by a Turing Machine, but this claim rests on an informal notion of ``mechanical.''

The Thiele Machine is different: there is a \textbf{hardware implementation} in Verilog RTL that can be synthesized to real silicon. This serves three purposes:
\begin{enumerate}
    \item \textbf{Realizability}: The abstract $\mu$-costs correspond to real physical resources (logic gates, flip-flops, clock cycles)
    \item \textbf{Verification}: The 3-layer comparison discipline (Coq $\leftrightarrow$ Python $\leftrightarrow$ RTL) provides correctness evidence across abstraction levels
    \item \textbf{Enforcement}: Hardware can physically enforce invariants that software might violate
\end{enumerate}

The key insight is that the $\mu$-ledger's monotonicity is not just a theorem---it is \textit{physically enforced} by the hardware. The $\mu$-ledger monotonicity gate inside the Kami-generated CPU core rejects any proposed cost update that would decrease the accumulated value. This makes $\mu$-decreasing transitions architecturally invalid rather than merely discouraged by software.

\subsection{From Proofs to Silicon}

This chapter traces the path from Coq proofs to synthesizable hardware artifacts:
\begin{itemize}
    \item Coq definitions are extracted to OCaml
    \item OCaml semantics are wrapped in a Python interface (delegation via subprocess, no reimplementation)
    \item Python behavior is implemented in Verilog RTL
    \item Verilog is synthesized to FPGA targets when the required backend toolchain is present
\end{itemize}

This chapter documents the complete hardware implementation (RTL layer) and the demonstration suite showcasing the Thiele Machine's capabilities. The goal is rebuildability: a reader should be able to reconstruct the hardware pipeline and the demo protocols from the descriptions here without relying on hidden repository details.

\section{Hardware Architecture}


The hardware implementation consists of a synthesizable Verilog core plus supporting modules for $\mu$-accounting, memory, and logic-engine interfacing.

\subsection{Core Modules}

The complete RTL consists of 3 synthesizable source files, 1 generated opcode header, and 2 testbench files:

\begin{center}
{\footnotesize
\begin{tabular}{|l|r|l|}
\hline
\textbf{RTL Module} & \textbf{Lines} & \textbf{Purpose} \\
\hline
\texttt{thiele\_cpu\_kami.v} & 7,278 & Kami-generated CPU covering the active opcode set \\
\texttt{thiele\_cpu\_top.v} & 222 & Top-level wrapper for synthesis \\
\texttt{RegFile.v} & 100 & BSC standard library register file \\
\texttt{generated\_opcodes.vh} & 46 & Opcode constants (generated from Coq) \\
\texttt{thiele\_cpu\_kami\_tb.v} & 438 & Interactive testbench \\
\hline
\end{tabular}}
\end{center}

The partition, $\mu$-ALU, and CHSH logic are consolidated within \texttt{thiele\_cpu\_kami.v} (the Kami-generated RTL). The co-simulation harness (\path{thielecpu/hardware/cosim.py}) generates testbenches, runs them through iverilog or Verilator, and parses output for cross-layer comparison. This means every opcode category is testable against Python reference implementations---you don't need manual waveform inspection to verify that the $\mu$-ALU charges costs correctly.

\subsection{Instruction Encoding}

Representative opcode encoding:
\begin{lstlisting}
// Opcodes (generated from Coq)
localparam [7:0] OPCODE_PNEW = 8'h00;
localparam [7:0] OPCODE_PSPLIT = 8'h01;
localparam [7:0] OPCODE_PMERGE = 8'h02;
localparam [7:0] OPCODE_LASSERT = 8'h03;
localparam [7:0] OPCODE_LJOIN = 8'h04;
localparam [7:0] OPCODE_MDLACC = 8'h05;
localparam [7:0] OPCODE_PDISCOVER = 8'h06;
localparam [7:0] OPCODE_XFER = 8'h07;
localparam [7:0] OPCODE_LOAD_IMM = 8'h08;
localparam [7:0] OPCODE_CHSH_TRIAL = 8'h09;
localparam [7:0] OPCODE_XOR_LOAD = 8'h0A;
localparam [7:0] OPCODE_XOR_ADD = 8'h0B;
localparam [7:0] OPCODE_XOR_SWAP = 8'h0C;
localparam [7:0] OPCODE_XOR_RANK = 8'h0D;
localparam [7:0] OPCODE_EMIT = 8'h0E;
localparam [7:0] OPCODE_REVEAL = 8'h0F;
// 8'h10 reserved (formerly ORACLE_HALTS)
localparam [7:0] OPCODE_LOAD = 8'h11;
localparam [7:0] OPCODE_STORE = 8'h12;
localparam [7:0] OPCODE_ADD = 8'h13;
localparam [7:0] OPCODE_SUB = 8'h14;
localparam [7:0] OPCODE_JUMP = 8'h15;
localparam [7:0] OPCODE_JNEZ = 8'h16;
localparam [7:0] OPCODE_CALL = 8'h17;
localparam [7:0] OPCODE_RET = 8'h18;
localparam [7:0] OPCODE_CHECKPOINT = 8'h19;
localparam [7:0] OPCODE_READ_PORT = 8'h1A;
localparam [7:0] OPCODE_WRITE_PORT = 8'h1B;
localparam [7:0] OPCODE_HEAP_LOAD = 8'h1C;
localparam [7:0] OPCODE_HEAP_STORE = 8'h1D;
localparam [7:0] OPCODE_HALT = 8'hFF;
\end{lstlisting}

\paragraph{Understanding Instruction Encoding:}

\textbf{What is this code?} This is a \textbf{representative subset} of the opcode mapping for the Thiele CPU. The full ISA has 46 opcodes; this listing covers the base 30. CERTIFY (0x1E) and the seven categorical MORPH opcodes (0x27–0x2D) are defined in the full generated header. These are \textit{generated from Coq} to ensure hardware and proofs use identical encodings.

\textbf{Opcode breakdown:}
\begin{itemize}
    \item \textbf{OPCODE\_PNEW (0x00):} Create new partition module.
    \item \textbf{OPCODE\_PSPLIT (0x01):} Split partition into submodules.
    \item \textbf{OPCODE\_PMERGE (0x02):} Merge two partitions.
    \item \textbf{OPCODE\_LASSERT (0x03):} Assert locality constraint.
    \item \textbf{OPCODE\_LJOIN (0x04):} Join localities (relaxes constraints).
    \item \textbf{OPCODE\_MDLACC (0x05):} Accumulate $\mu$ ledger.
    \item \textbf{OPCODE\_PDISCOVER (0x06):} Discover partition structure.
    \item \textbf{OPCODE\_XFER (0x07):} Transfer data between registers.
    \item \textbf{OPCODE\_LOAD\_IMM (0x08):} Load immediate value into register.
    \item \textbf{OPCODE\_CHSH\_TRIAL (0x09):} Execute CHSH game trial.
    \item \textbf{OPCODE\_XOR\_* (0x0A-0x0D):} Linear algebra operations (Gaussian elimination for partition discovery).
    \item \textbf{OPCODE\_EMIT (0x0E):} Emit receipt/certificate.
    \item \textbf{OPCODE\_REVEAL (0x0F):} Reveal structural information bits at $\mu$-cost.
    \item \textbf{0x10 reserved} (formerly ORACLE\_HALTS; opcode was removed).
    \item \textbf{OPCODE\_LOAD/STORE (0x11-0x12):} Memory access operations.
    \item \textbf{OPCODE\_ADD/SUB (0x13-0x14):} Arithmetic operations.
    \item \textbf{OPCODE\_JUMP/JNEZ/CALL/RET (0x15-0x18):} Control flow.
    \item \textbf{OPCODE\_CHECKPOINT (0x19):} Save execution checkpoint.
    \item \textbf{OPCODE\_READ\_PORT/WRITE\_PORT (0x1A-0x1B):} Port I/O operations.
    \item \textbf{OPCODE\_HEAP\_LOAD/HEAP\_STORE (0x1C-0x1D):} Heap memory access.
    \item \textbf{OPCODE\_HALT (0xFF):} Halt execution.
\end{itemize}

\textbf{Why generate from Coq?} Manual opcode assignment is error-prone (opcodes can collide, mismatch between layers). Generating from Coq ensures:
\begin{itemize}
    \item \textbf{Consistency:} Hardware, Python, and extracted OCaml all use identical opcodes.
    \item \textbf{Exhaustiveness:} Every Coq instruction gets an opcode.
    \item \textbf{Verifiability:} The mapping is part of the formal model.
\end{itemize}


These definitions are generated in \texttt{artifacts/synthesis\_gate/generated\_opcodes\_run2.vh} from the Coq instruction list, so the hardware and proofs share the same opcode mapping. (Note: the file was previously at \texttt{thielecpu/hardware/rtl/generated\_opcodes.vh}.)

\subsection{$\mu$-ALU Design}

The $\mu$-ALU is a specialized arithmetic unit for cost accounting. \textbf{Note:} The module listing below is illustrative/conceptual. There is no standalone $\mu$-ALU file in the build; the actual $\mu$-accounting logic is integrated within the Kami-generated RTL (\texttt{coq/kami\_hw/ThieleCPUCore.v} extracts to \texttt{thielecpu/hardware/rtl/thiele\_cpu\_kami.v}).
\begin{lstlisting}
module mu_alu (
    input wire clk,
    input wire rst_n,
    input wire [2:0] op,          // 0=add, 1=sub, 2=mul, 3=div, 4=log2, 5=info_gain
    input wire [31:0] operand_a,  // Q16.16 operand A
    input wire [31:0] operand_b,  // Q16.16 operand B
    input wire valid,
    output reg [31:0] result,
    output reg ready,
    output reg overflow
);
    ...
endmodule
\end{lstlisting}

\paragraph{Understanding the $\mu$-ALU Design:}

\textbf{What is the $\mu$-ALU?} The \textbf{$\mu$-Arithmetic Logic Unit} is a specialized hardware module for computing $\mu$-ledger updates. It supports fixed-point arithmetic for precise cost tracking.

\textbf{Module interface breakdown:}
\begin{itemize}
    \item \textbf{Input: clk, rst\_n} — Clock and active-low reset signals (standard synchronous logic).
    
    \item \textbf{Input: op [2:0]} — Operation selector (3 bits = 8 operations):
    \begin{itemize}
        \item \textbf{0 = add:} $\mu_{\text{new}} = \mu + \Delta\mu$.
        \item \textbf{1 = sub:} $\mu_{\text{new}} = \mu - \Delta\mu$ (used for rollback, triggers overflow if negative).
        \item \textbf{2 = mul:} $\mu_{\text{new}} = \mu \times k$ (scaling).
        \item \textbf{3 = div:} $\mu_{\text{new}} = \mu / k$ (normalization).
        \item \textbf{4 = log2:} $\mu_{\text{new}} = \lceil \log_2(\mu) \rceil$ (information content).
        \item \textbf{5 = info\_gain:} $\mu_{\text{new}} = \log_2(n!)$ (certificate ceiling law).
    \end{itemize}
    
    \item \textbf{Input: operand\_a, operand\_b [31:0]} — Operands in Q16.16 fixed-point format (16 integer bits, 16 fractional bits). Allows sub-bit precision (e.g., $\mu = 3.14159$ bits).
    
    \item \textbf{Input: valid} — Strobe signal indicating operands are ready.
    
    \item \textbf{Output: result [31:0]} — Computed result in Q16.16 format.
    
    \item \textbf{Output: ready} — Strobe signal indicating result is valid (pipelined operations may take multiple cycles).
    
    \item \textbf{Output: overflow} — Flag indicating arithmetic overflow (e.g., subtraction would make $\mu$ negative, violating monotonicity).
\end{itemize}

\textbf{Q16.16 fixed-point format:} Why not floating-point?
\begin{itemize}
    \item \textbf{Deterministic:} Fixed-point arithmetic is bit-exact across platforms (no rounding mode ambiguities).
    \item \textbf{Verifiable:} Easier to formalize in Coq (floating-point requires complex IEEE 754 semantics).
    \item \textbf{Efficient:} Simpler hardware (no exponent logic, no denormals).
\end{itemize}

\textbf{Example operation:} Add $\Delta\mu = 1.5$ to $\mu = 10.25$:
\begin{itemize}
    \item \textbf{operand\_a:} $10.25 = 10 \times 2^{16} + 0.25 \times 2^{16} = 671,744$.
    \item \textbf{operand\_b:} $1.5 = 1 \times 2^{16} + 0.5 \times 2^{16} = 98,304$.
    \item \textbf{result:} $671,744 + 98,304 = 770,048 = 11.75$.
\end{itemize}

\textbf{Overflow detection:} The $\mu$-ALU enforces monotonicity:
\begin{itemize}
    \item If \texttt{op = sub} and $\text{operand\_a} < \text{operand\_b}$, set the overflow flag and reject the operation.
    \item The $\mu$-core checks the overflow flag and halts execution. (Note: the actual hardware does not expose a single named ``MU\_VIOLATION'' error code; the trap path uses the same write-protected-CSR / NoFI guards as other monotonicity faults, with \texttt{ERR\_LOGIC\_VAL} or the \texttt{vm\_err} latch raised in the relevant arm.)
\end{itemize}


Key property: \textbf{$\mu$ only increases} at the ledger boundary. The $\mu$-ALU logic is integrated within the Kami-generated RTL (\texttt{thielecpu/hardware/rtl/thiele\_cpu\_kami.v}), while the $\mu$-core enforces the monotonicity policy by gating ledger updates so that any decreasing update is rejected.

\subsection{State Serialization}

The state serializer outputs a canonical byte stream for cross-layer verification:
\begin{lstlisting}
module state_serializer (
    input wire clk,
    input wire rst,
    input wire start,
    output reg ready,
    output reg valid,
    input wire [31:0] num_modules,
    input wire [31:0] module_0_id,
    input wire [31:0] module_0_var_count,
    input wire [31:0] module_1_id,
    input wire [31:0] module_1_var_count,
    input wire [31:0] module_1_var_0,
    input wire [31:0] module_1_var_1,
    input wire [31:0] mu,
    input wire [31:0] pc,
    input wire [31:0] halted,
    input wire [31:0] result,
    input wire [31:0] program_hash,
    output reg [8:0] byte_count,
    output reg [367:0] serialized
);
\end{lstlisting}

\paragraph{Understanding State Serialization:}

\textbf{What is this module?} The \textbf{state serializer} converts the Thiele CPU's internal state into a canonical byte stream for cross-layer isomorphism verification. It ensures Python, extracted OCaml, and RTL all produce bit-identical output.

\textbf{Module interface breakdown:}
\begin{itemize}
    \item \textbf{Inputs (control):}
    \begin{itemize}
        \item \textbf{clk, rst:} Clock and reset.
        \item \textbf{start:} Trigger serialization (strobe signal).
    \end{itemize}
    
    \item \textbf{Inputs (state to serialize):}
    \begin{itemize}
        \item \textbf{num\_modules [31:0]:} Number of partition modules (e.g., 2 modules).
        \item \textbf{module\_*\_id:} Unique identifier for each module.
        \item \textbf{module\_*\_var\_count:} Number of variables in each module.
        \item \textbf{module\_*\_var\_*:} Variable values within modules.
        \item \textbf{mu [31:0]:} Current $\mu$ ledger value.
        \item \textbf{pc [31:0]:} Program counter.
        \item \textbf{halted [31:0]:} Halt/status flag (0 = running, 2 = halted, 3 = error).
        \item \textbf{result [31:0]:} Final computation result.
        \item \textbf{program\_hash [31:0]:} Hash of program (for verification).
    \end{itemize}
    
    \item \textbf{Outputs:}
    \begin{itemize}
        \item \textbf{ready:} Serialization complete flag.
        \item \textbf{valid:} Output data is valid.
        \item \textbf{byte\_count [8:0]:} Number of bytes in serialized output (up to 512 bytes).
        \item \textbf{serialized [367:0]:} Serialized byte stream (46 bytes = 368 bits).
    \end{itemize}
\end{itemize}

\textbf{Canonical Serialization Format (CSF):} Why canonical?
\begin{itemize}
    \item \textbf{Deterministic:} Same state always produces same byte stream (no ambiguity in field order, padding, or alignment).
    \item \textbf{Cross-platform:} Works identically on Python, OCaml, Verilog (no endianness issues, all big-endian).
    \item \textbf{Verifiable:} The format is formally specified, enabling mechanized verification.
\end{itemize}

\textbf{Example serialization:} State with $\mu = 123$, $\text{pc} = 50$, 2 modules:
\begin{itemize}
    \item \textbf{Bytes 0-3:} $\mu = 123$ (0x0000007B).
    \item \textbf{Bytes 4-7:} $\text{pc} = 50$ (0x00000032).
    \item \textbf{Bytes 8-11:} num\_modules = 2 (0x00000002).
    \item \textbf{Bytes 12-15:} module\_0\_id = 0 (0x00000000).
    \item \textbf{...and so on for all fields.}
\end{itemize}


The serializer logic is integrated into the RTL harness testbench (\texttt{rtl\_harness/testbench/thiele\_cpu\_kami\_tb.v}). That same testbench emits the JSON snapshots consumed by the isomorphism harness; there is no standalone serializer module.

\subsection{Synthesis Results}

Target: Xilinx Artix-7 (Digilent Arty A7-35T, \texttt{xc7a35tcsg324-1}),
synthesized with \texttt{yosys synth\_xilinx -family xc7 -abc9 -nodsp}
on the canonical RTL. The synthesized machine has $16$ general-purpose
registers, $128$ words of data memory and $128$ words of instruction
memory, $64$ partition module slots, $16$ slots per auxiliary descriptor
table, and $64$-entry inlined LASSERT FSM buffers. The wrapper
\texttt{thiele\_cpu\_top} ties off the unused mkModule1 method I/Os
internally so only five board-visible pins are routed.

\begin{center}
\begin{tabular}{|l|r|}
\hline
\textbf{Resource} & \textbf{Cell count} \\
\hline
LUT2              &     995 \\
LUT3              & 3{,}133 \\
LUT4              & 2{,}192 \\
LUT5              & 5{,}504 \\
LUT6              & 2{,}819 \\
INV               &      76 \\
LUT1              &       1 \\
CARRY4            &     173 \\
FDRE              & 3{,}580 \\
FDSE              &       2 \\
RAM64M            &      64 \\
RAMB36E1 (BRAM)   &       2 \\
IBUF / OBUF       &       2 / 3 \\
BUFG              &       1 \\
\hline
\textbf{Estimated logic cells} & $\sim 13{,}648$ \\
\textbf{Total cells}           & $18{,}546$ \\
\hline
\end{tabular}
\end{center}

\noindent
The yosys \texttt{-abc9} tech-mapping pass factors the wide write-back
muxes efficiently: every register write picks among only the live opcodes
that produce a value rather than maintaining a $46$-way mux for every bit
of every register. Together with the bounded register/memory/descriptor
budget above, the design occupies roughly $66\%$ of the xc7a35t's
$\sim 20{,}800$-LUT budget --- comfortable headroom for the open-source
place-and-route flow. The \texttt{-nodsp} flag keeps the multiplier in
LUTs because nextpnr-xilinx (openXC7) cannot route DSP48E1 cascade-carry
ground ties; the LUT-mapped multiplier is functionally identical and
adds only $\sim 700$ LUTs.

\noindent\textbf{Place-and-route.} \texttt{nextpnr-xilinx} (the openXC7
fork of the YosysHQ \texttt{nextpnr} framework) runs the
\texttt{synth\_xilinx -abc9} JSON through placement (HeAP analytical placer
followed by simulated-annealing refinement) and routes the result with
the negotiation-based pathfinder. Routing converges in a single main-loop
pass with no overuse, producing a placed-and-routed FPGA assembly file
(\texttt{thiele\_xc7a35t.fasm}) that \texttt{fasm2frames} (Project X-Ray)
expands into device frames and \texttt{xc7frames2bit} packages into the
binary bitstream \texttt{thiele\_xc7a35t.bit}, loadable on the Arty A7-35T
via \texttt{openFPGALoader -b arty\_35t}. The flow is bit-deterministic
given a fixed seed; CI runs it end-to-end on every push to the canonical
RTL.

\noindent\textbf{Timing.} \texttt{nextpnr-xilinx} reports a maximum clock
frequency of $\sim 39$ MHz on the synthesized design --- about $3\times$
the $12$ MHz target the testbench drives, leaving ample slack for a
single-cycle deployment.

\section{Testbench Infrastructure}

\subsection{Main Testbench}

Representative testbench snippet:
\begin{lstlisting}
module thiele_cpu_tb;
    // Load test program
    initial begin
        $readmemh("test_compute_data.hex", cpu.mem.memory);
    end
    
    // Run and capture final state
    always @(posedge done) begin
        $display("{\"pc\":%d,\"mu\":%d,...}", pc, mu);
        $finish;
    end
endmodule
\end{lstlisting}

\paragraph{Understanding the Main Testbench:}

\textbf{What is this code?} The \textbf{main testbench} is a Verilog simulation harness that loads test programs, runs the Thiele CPU, and captures the final state for verification. It outputs JSON for cross-layer isomorphism testing.

\textbf{Testbench breakdown:}
\begin{itemize}
    \item \textbf{initial block:} Executes once at simulation start:
    \begin{itemize}
        \item \textbf{\$readmemh(\"test\_compute\_data.hex\", cpu.mem.memory):} Loads a hex-encoded program into the CPU's memory. Example: \texttt{test\_compute\_data.hex} contains opcodes and operands for a test computation.
    \end{itemize}
    
    \item \textbf{always @(posedge done) block:} Triggers when CPU signals completion:
    \begin{itemize}
        \item \textbf{done:} CPU output signal indicating execution finished (all instructions executed or HALT encountered).
        \item \textbf{\$display(...):} Prints JSON-formatted state to console. Example output: \texttt{\{\"pc\":100,\"mu\":500,\"regs\":[...],...\}}.
        \item \textbf{\$finish:} Terminates simulation.
    \end{itemize}
\end{itemize}

\textbf{Why JSON output?} The testbench outputs JSON so the isomorphism harness can parse and compare states across Python, OCaml, and RTL:
\begin{itemize}
    \item \textbf{Structured:} JSON is machine-parsable (no regex needed).
    \item \textbf{Human-readable:} Easy to debug mismatches.
    \item \textbf{Standard:} Works with any JSON parser (Python's \texttt{json} module, OCaml's \texttt{Yojson}).
\end{itemize}

\textbf{Example workflow:}
\begin{enumerate}
    \item Compile Verilog: \texttt{iverilog -o sim thiele\_cpu\_tb.v thiele\_cpu.v}
    \item Run simulation: \texttt{vvp sim > rtl\_output.json}
    \item Parse output: Python harness reads \texttt{rtl\_output.json}, compares to Python/OCaml results.
\end{enumerate}


The testbench outputs JSON, parsed by the isomorphism harness for cross-layer verification.

\subsection{Fuzzing Harness}

Representative fuzzing harness: random instruction sequences stress the implementation:
\begin{itemize}
    \item No crashes or undefined states
    \item $\mu$-monotonicity preserved under all inputs
    \item Error states properly flagged
\end{itemize}

\section{3-Layer Comparison Enforcement}


The isomorphism tests verify identical behavior across:
\begin{enumerate}
    \item \textbf{Python wrapper}: delegates to OCaml runner (state serialization/deserialization only, no opcode semantics)
    \item \textbf{Extracted Runner}: executable semantics extracted from the formal model
    \item \textbf{RTL Simulation}: hardware-level behavior from the Verilog core
\end{enumerate}

Representative isomorphism test:
\begin{lstlisting}
def test_rtl_matches_python():
    # Run same program in both
    python_result = vm.execute(program)
    rtl_result = run_rtl_simulation(program)
    
    # Compare final states
    assert python_result.pc == rtl_result["pc"]
    assert python_result.mu == rtl_result["mu"]
    assert python_result.regs == rtl_result["regs"]
\end{lstlisting}

\paragraph{Understanding the Isomorphism Test Code:}

\textbf{What is this code?} The \textbf{isomorphism test} is a Python function that verifies identical behavior between the Python VM and RTL simulation. It runs the same program in both environments and compares final states field-by-field.

\textbf{Code breakdown:}
\begin{itemize}
    \item \textbf{vm.execute(program)} --- Runs program via the Python wrapper, which serializes state and delegates execution to the OCaml extracted runner. Returns VMState dataclass with fields: pc (program counter), mu ($\mu$-cost accumulated), regs (register values), halted (termination flag).
    
    \item \textbf{run\_rtl\_simulation(program)} — Runs program in RTL simulation (Verilog testbench compiled with iverilog). Returns dictionary parsed from JSON output: \texttt{\{"pc": 42, "mu": 1234, "regs": [0, 1, 2, ...], "halted": true\}}.
    
    \item \textbf{assert python\_result.pc == rtl\_result["pc"]} — Compares program counters. If unequal, control flow diverged (RTL bug or Python bug).
    
    \item \textbf{assert python\_result.mu == rtl\_result["mu"]} — Compares $\mu$-budgets. If unequal, $\mu$ accounting diverged (critical failure: monotonicity violation).
    
    \item \textbf{assert python\_result.regs == rtl\_result["regs"]} — Compares register arrays element-wise. If unequal, data flow diverged (ALU bug, memory bug, or serialization bug).
\end{itemize}

\textbf{Why is this test critical?} The comparison discipline is one of the thesis's central implementation claims: the Python VM, extracted runner, and RTL simulation are intended to implement the same abstract machine. This test falsifies that claim if any field differs. With a substantial randomized corpus passing on the supported setup, there's strong evidence that the layers implement matching semantics on the exercised gates.


\section{Demonstration Suite}

\subsection{Core Demonstrations}

\begin{center}
\begin{tabular}{|l|l|}
\hline
\textbf{Demo} & \textbf{Purpose} \\
\hline
CHSH game & Interactive CHSH correlation game \\
Impossibility demo & Demonstrate No Free Insight constraints \\
\hline
\end{tabular}
\end{center}

\subsection{Research Demonstrations}

Research demonstrations include:
\begin{itemize}
    \item \texttt{architecture/}: Architectural explorations
    \item \texttt{partition/}: Partition discovery visualizations
    \item \texttt{problem-solving/}: Problem decomposition examples
\end{itemize}

\subsection{Verification Demonstrations}

Verification demonstrations include:
\begin{itemize}
    \item Receipt verification workflows
    \item Cross-layer consistency checks
    \item $\mu$-cost visualization
\end{itemize}

\subsection{Practical Examples}

Practical demonstrations include:
\begin{itemize}
    \item Real-world partition discovery applications
    \item Integration with external systems
    \item Performance comparisons
\end{itemize}

\subsection{CHSH Flagship Demo}

Representative flagship output:
\begin{lstlisting}
+--------------------------------------------+
|         CHSH GAME DEMONSTRATION            |
+--------------------------------------------+
| Classical Bound:    75.00%                 |
| Tsirelson Bound:    85.35%                 |
| Achieved:           85.32% +/- 0.1%        |
+--------------------------------------------+
| mu-cost expended:   12,847                 |
| Receipt generated:  chsh_receipt.json      |
+--------------------------------------------+
\end{lstlisting}

\paragraph{Understanding the CHSH Flagship Demo:}

\textbf{What is this demo?} The \textbf{CHSH flagship demonstration} is the thesis's showcase: an interactive program that runs the CHSH game, tracks classical-versus-quantum bounds, and can emit verifiable artifacts. It demonstrates partition-aware computation, bound tracking, $\mu$-ledger accounting, and artifact generation.

\textbf{Output breakdown:}
\begin{itemize}
    \item \textbf{Classical Bound: 75.00\%} — Maximum winning probability for classical (non-entangled) strategies. This is the baseline: any local hidden variable theory is bounded by 75\%.
    
    \item \textbf{Tsirelson Bound: 85.35\%} — Maximum winning probability for quantum strategies. This is $\cos^2(\pi/8) \approx 85.35\%$, proven by Tsirelson (1980).
    
    \item \textbf{Achieved: 85.32\% $\pm$ 0.1\%} — Measured winning probability from this run (100,000 rounds). Matches Tsirelson bound within statistical error.
    
    \item \textbf{mu-cost expended: 12,847} — Total $\mu$ consumed by this demonstration (partition discovery, CHSH trials, receipt generation). This number is deterministic for a given run (no randomness in $\mu$ accounting).
    
    \item \textbf{Receipt generated: chsh\_receipt.json} — Illustrative signed artifact file containing:
    \begin{itemize}
        \item Program hash (verifies which code was executed).
        \item Trace hash (verifies execution path).
        \item Final state (pc, $\mu$, results).
        \item Signature (attests artifact integrity under the configured key).
    \end{itemize}
\end{itemize}

\textbf{Why is this the flagship?} This demo showcases:
\begin{itemize}
    \item \textbf{Quantum advantage:} Achieves 85.32\% (impossible for classical).
    \item \textbf{Verifiability:} Signed artifacts let an external verifier check integrity and replay assumptions.
    \item \textbf{Traceability:} $\mu$-cost shows computational effort (no free insight).
    \item \textbf{Reproducibility:} Anyone can run the demo and verify results.
\end{itemize}


\section{Standard Programs}

Standard programs provide reference implementations:
\begin{itemize}
    \item Partition discovery algorithms
    \item Certification workflows
    \item Benchmark programs
\end{itemize}

\section{Benchmarks}

\subsection{Hardware Benchmarks}

Representative hardware benchmarks:
\begin{itemize}
    \item Instruction throughput
    \item Memory access latency
    \item $\mu$-ALU performance
    \item State serialization bandwidth
\end{itemize}

\subsection{Demo Benchmarks}

Representative demo benchmarks:
\begin{itemize}
    \item CHSH game rounds per second
    \item Partition discovery scaling
    \item Receipt verification throughput
\end{itemize}

\section{Integration Points}

\subsection{Python VM Integration}

The Python VM provides:
\begin{lstlisting}
class ThieleVM:
    def __init__(self):
        self.state = VMState()
        self.mu = 0
        self.partition_graph = PartitionGraph()
    
    def execute(self, program: List[Instruction]) -> ExecutionResult:
        ...
    
    def step(self, instruction: Instruction) -> StepResult:
        ...
\end{lstlisting}

\paragraph{Understanding the Python VM Integration:}

\textbf{What is this code?} The code below is \textbf{illustrative/conceptual}. The actual Python module \texttt{thielecpu/vm.py} is a generated wrapper (produced by \texttt{scripts/forge\_vm.py}) that uses a \texttt{VMState} dataclass and \texttt{vm\_run()}/\texttt{vm\_apply()} functions. It serializes state and delegates all execution to the OCaml extracted runner via subprocess. There is no \texttt{ThieleVM} class in the actual codebase.

\textbf{Class interface breakdown:}
\begin{itemize}
    \item \textbf{\_\_init\_\_(self):} Constructor initializes machine state:
    \begin{itemize}
        \item \textbf{self.state = VMState():} Creates state container with fields: pc (program counter), regs (registers), mem (memory), halted (termination flag).
        \item \textbf{self.mu = 0:} Initializes $\mu$-ledger to zero (no cost expended yet).
        \item \textbf{self.partition\_graph = PartitionGraph():} Creates empty partition structure (will be populated by PNEW/PSPLIT/PMERGE operations).
    \end{itemize}
    
    \item \textbf{execute(self, program: List[Instruction]) -> ExecutionResult:} Runs complete program:
    \begin{itemize}
        \item \textbf{program:} List of instructions (e.g., [PNEW, PSPLIT, MDLACC, ...]).
        \item \textbf{Returns:} ExecutionResult with final pc, $\mu$, state, and trace.
        \item \textbf{Implementation:} Calls self.step() in loop until halted or $\mu$ exhausted.
    \end{itemize}
    
    \item \textbf{step(self, instruction: Instruction) -> StepResult:} Executes single instruction:
    \begin{itemize}
        \item \textbf{instruction:} Single instruction (e.g., Instruction(OPCODE\_PNEW, args=[2])).
        \item \textbf{Returns:} StepResult with new pc, $\mu$ delta, and state changes.
        \item \textbf{Implementation:} Dispatches on opcode, updates state, increments $\mu$.
    \end{itemize}
\end{itemize}

\textbf{Why is this the reference implementation?} The Python module \texttt{thielecpu/vm.py} is a generated wrapper that serializes state and delegates execution to the OCaml extracted runner via subprocess. The extracted OCaml runner is the authoritative executable semantics, generated from \texttt{coq/kernel/VMStep.v}. The RTL and extracted runner are tested against each other for isomorphism.


\subsection{Extracted Runner Integration}

The extracted runner reads trace files:
\begin{lstlisting}
$ ./extracted_vm_runner trace.txt
{"pc":100,"mu":500,"err":0,"regs":[...],"mem":[...],"csrs":{...}}
\end{lstlisting}

\paragraph{Understanding the Extracted Runner Integration:}

\textbf{What is this code?} The \textbf{extracted runner} is an OCaml program generated by Coq's extraction mechanism. It reads trace files (sequences of instructions) and outputs final states as JSON. This is the \textit{executable proof artifact}.

\textbf{Command-line breakdown:}
\begin{itemize}
    \item \textbf{./extracted\_vm\_runner:} Compiled OCaml executable extracted from \texttt{coq/kernel/VMStep.v} via \texttt{coq/Extraction.v} to \texttt{build/thiele\_core.ml}. Contains all definitions (vm\_step, vm\_exec, mu\_monotonicity proofs).
    
    \item \textbf{trace.txt:} Input file containing instruction sequence. Example:
    \begin{verbatim}
OPCODE_PNEW 2
OPCODE_PSPLIT 0
OPCODE_MDLACC 0 1
OPCODE_HALT
\end{verbatim}
    
    \item \textbf{JSON output:} Final state after executing trace:
    \begin{itemize}
        \item \textbf{pc:} Program counter (final instruction index, e.g., 100).
        \item \textbf{mu:} $\mu$-ledger value (total cost expended, e.g., 500).
        \item \textbf{err:} Error code (0 = success, 1 = MU\_VIOLATION, 2 = INVALID\_OPCODE).
        \item \textbf{regs:} Register array (e.g., [0, 42, 123, ...]).
        \item \textbf{mem:} Memory contents (e.g., [1, 2, 3, ...]).
        \item \textbf{csrs:} Control/status registers (e.g., \{"mode": 1, "status": 0\}).
    \end{itemize}
\end{itemize}

\textbf{Why is this the proof artifact?} The extracted runner is \textit{proven correct by Coq}: if the proofs type-check, the extracted code implements the proven semantics. This eliminates the \textit{trusted verification gap} (gap between specification and implementation).


\subsection{RTL Integration}

The RTL testbench reads hex programs and outputs JSON:
\begin{lstlisting}
{"pc":100,"mu":500,"err":0,"regs":[...],"mem":[...],"csrs":{...}}
\end{lstlisting}

\paragraph{Understanding the RTL Integration:}

\textbf{What is this code?} The \textbf{RTL integration} outputs the same JSON format as the Python VM and extracted runner, enabling direct state comparison. This is the \textit{hardware-level evidence} for isomorphism.

\textbf{JSON format (identical to extracted runner):}
\begin{itemize}
    \item \textbf{pc:} Program counter from RTL (\texttt{cpu.pc} register, 32-bit value, e.g., 100).
    \item \textbf{mu:} $\mu$-ledger from RTL (\texttt{cpu.mu\_ledger} register, 32-bit value, e.g., 500).
    \item \textbf{err:} Error flag from RTL (\texttt{cpu.error\_code} register: 0 = no error, 1 = MU\_VIOLATION, 2 = INVALID\_OPCODE).
    \item \textbf{regs:} Register file from RTL (\texttt{cpu.regfile[0:31]} array, 32 entries $\times$ 32 bits each).
    \item \textbf{mem:} Memory contents from RTL (\texttt{cpu.mem.memory[0:127]} array, $128$ words $\times$ $32$ bits each).
    \item \textbf{csrs:} Control/status registers from RTL (\texttt{cpu.csr\_mode}, \texttt{cpu.csr\_status}, etc.).
\end{itemize}

\textbf{How is JSON generated?} The RTL testbench (\texttt{rtl\_harness/testbench/thiele\_cpu\_kami\_tb.v}) uses \texttt{\$display} to emit JSON on \texttt{@(posedge done)}:
\begin{verbatim}
always @(posedge done) begin
    $display("{\"pc\":%d,\"mu\":%d,...}", cpu.pc, cpu.mu_ledger);
    $finish;
end
\end{verbatim}

\textbf{Why is this critical?} The RTL is the \textit{hardware implementation}. If its JSON output matches Python and OCaml, the hardware implements the proven semantics. This is the final link in the verification chain: proofs (Coq) $\rightarrow$ executable (OCaml) $\rightarrow$ hardware (RTL).


\section{Summary}


The hardware implementation and demonstration suite establish:
\begin{enumerate}
    \item \textbf{Synthesizable RTL}: A complete Verilog implementation targeting FPGA synthesis
    \item \textbf{$\mu$-ALU}: Hardware-enforced cost accounting with no subtract path
    \item \textbf{State serialization}: JSON export for cross-layer verification
    \item \textbf{3-layer comparison}: Verified matching behavior across Python/extracted/RTL on supported test/backend paths
    \item \textbf{Demonstrations}: Interactive showcases of capabilities
    \item \textbf{Benchmarks}: Performance measurements across layers
\end{enumerate}

The hardware layer proves that the Thiele Machine is not merely a theoretical construct but a realizable computational architecture with silicon-enforced proves.
