\documentclass[9pt, a4paper, oneside, twocolumn]{extreport}

% ============================================================================
% PACKAGES
% ============================================================================
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{times}
\usepackage{graphicx}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{geometry}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{mdframed}
\usepackage{booktabs}
\usepackage{fancyhdr}
\usepackage{setspace}
\usepackage{enumitem}
\usepackage{tocloft}
\usepackage[bookmarks=true,colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{xurl}
\usepackage{microtype}

% TikZ for diagrams
\usepackage{tikz}
\usetikzlibrary{shapes, arrows.meta, positioning, calc, fit, backgrounds, decorations.pathreplacing, decorations.pathmorphing}

% PGFPlots for plots
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}

% Tcolorbox for colored boxes
\usepackage{tcolorbox}
\tcbuselibrary{skins, breakable}

% Tcolorbox styling for two-column layout
\tcbset{
    thesisbox/.style={
        enhanced,
        breakable,
        pad at break*=4pt,
        width=\linewidth,
        boxrule=0.6pt,
        arc=0pt,
        colback=white,
        colframe=black!70,
        left=5pt,
        right=5pt,
        top=5pt,
        bottom=5pt,
        boxsep=3pt,
        before skip=7pt,
        after skip=7pt,
        fontupper=\footnotesize,
        fonttitle=\bfseries,
        coltitle=white,
        boxed title style={
            colback=black!70,
            colframe=black!70,
            boxrule=0.6pt,
            arc=0pt,
            left=4pt,
            right=4pt,
            top=2pt,
            bottom=2pt
        }
    }
}

% Float placement controls
\usepackage{float}
\usepackage{placeins}

% Better float placement parameters
\renewcommand{\topfraction}{0.9}
\renewcommand{\bottomfraction}{0.8}
\renewcommand{\textfraction}{0.1}
\renewcommand{\floatpagefraction}{0.7}
\setcounter{topnumber}{3}
\setcounter{bottomnumber}{3}
\setcounter{totalnumber}{4}

% Ensure figures don't exceed text width
\setkeys{Gin}{width=\linewidth,keepaspectratio}

% Allow breaking in texttt
\usepackage[htt]{hyphenat}

% Handle Unicode characters
\DeclareUnicodeCharacter{03BC}{\ensuremath{\mu}}
\DeclareUnicodeCharacter{2014}{-}

% Table of contents depth (show up to subsections)
\setcounter{tocdepth}{2}
\setcounter{secnumdepth}{3}

% ============================================================================
% PAGE GEOMETRY
% ============================================================================
\geometry{
    top=0.45in,
    bottom=0.45in,
    left=0.6in,
    right=0.6in
}

% Fix headheight warning
\setlength{\headheight}{14.5pt}

% Line spacing
\linespread{0.92}
\setlength{\parskip}{2pt}
\setlength{\parindent}{1em}
\setlist{itemsep=1pt, topsep=2pt, parsep=0pt, partopsep=0pt}
\setlength{\columnsep}{0.2in}
\raggedbottom

% ============================================================================
% HEADERS AND FOOTERS
% ============================================================================
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\leftmark}
\fancyhead[R]{\thepage}
\renewcommand{\headrulewidth}{0.4pt}

% ============================================================================
% THEOREM ENVIRONMENTS
% ============================================================================
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Fix: add vertical space after theorem header so lstlisting frames
% don't overlap with the theorem title text
\makeatletter
\let\orig@theorem\theorem
\def\theorem{\orig@theorem\vspace{2pt}}
\let\orig@lemma\lemma
\def\lemma{\orig@lemma\vspace{2pt}}
\let\orig@corollary\corollary
\def\corollary{\orig@corollary\vspace{2pt}}
\makeatother

% ============================================================================
% CODE LISTINGS
% ============================================================================
\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},   
    commentstyle=\color{codegreen},
    keywordstyle=\color{blue},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily\scriptsize,
    columns=fullflexible,
    breakatwhitespace=true,         
    breaklines=true,                 
    captionpos=b,                    
    keepspaces=true,                 
    numbers=none,
    showspaces=false,                
    showstringspaces=false,
    showtabs=false,                  
    tabsize=2,
    frame=single,
    framesep=2pt,
    framexleftmargin=1pt,
    framexrightmargin=1pt,
    framerule=0.5pt,
    rulecolor=\color{codegray},
    aboveskip=1em,
    belowskip=1em,
    xleftmargin=0pt,
    xrightmargin=0pt
}
\lstset{style=mystyle}

% Prevent page breaks in short listings
\lstset{
    postbreak=\mbox{\textcolor{red}{$\hookrightarrow$}\space},
    escapeinside={(*@}{@*)},
    float=h
}

% ============================================================================
% TABLE OF CONTENTS FORMATTING
% ============================================================================
\renewcommand{\cfttoctitlefont}{\hfill\Large\bfseries}
\renewcommand{\cftaftertoctitle}{\hfill}
\setlength{\cftbeforechapskip}{1em}

% ============================================================================
% TITLE
% ============================================================================
\title{
    \textbf{\Huge The Thiele Machine}\\
    \Large Structural Entitlement, Accountable Computation, and No Free Insight\\
    \large A Formal Model Built by Asking Questions
}
\author{
    \textbf{Devon Thiele}\\
    \normalsize April 2026
}
\date{}

% ============================================================================
% DOCUMENT
% ============================================================================
\begin{document}

% Title page
\twocolumn[
\maketitle
\thispagestyle{empty}
]

% Abstract
\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

What if discovering that a problem decomposes---that variables are independent, that a graph has two components---were not free?

The Turing machine is a complete model of computable functions. It is provably silent on computational epistemology. A Turing machine computes whether a list is sorted; it issues no unforgeable receipt for what that knowing cost. The \textbf{Thiele Machine} is the formal system that makes computation speak about its own epistemic cost. The \textbf{$\mu$-ledger} is the receipt: necessary, unique, and inaccessible to any classical observer.

I'm a car salesman. I have no formal training in computer science, mathematics, or physics. I built this through LLM-directed development: I specify the ideas, the invariants, and the falsification conditions, then direct AI tools to implement them. The proofs compile or they don't. They don't care who wrote them.

\textbf{What is proven} (in Coq, zero admits, zero global axioms in the active audited tree; two unavoidable gaps are named explicitly as Section Variables rather than hidden as global postulates: \texttt{A\_QM} in \path{coq/PhysicsConditionalClosure.v} (the physical claim that honest quantum experiments produce PSD correlations) and \texttt{bsc\_kami\_compilation\_trusted} in \path{coq/kernel/VerilogRTLCorrespondence.v} (the BSC pretty-printer and Bluespec$\to$Verilog compilation step); standard library imports remain):
\begin{itemize}
    \item \textbf{No Free Insight}: A trace that produces a certified final state must contain at least one cost-paying instruction (\path{coq/kernel/NoFreeInsight.v}, theorem \texttt{no\_free\_insight\_general}). The quantitative companion $\Delta\mu \ge \log_2(\Omega) - \log_2(\Omega')$ is the Shannon-bridge bound in \path{coq/kernel/MuShannonBridge.v}.
    \item \textbf{$\mu$-Conservation}: The ledger grows monotonically; every instruction adds exactly its declared cost, no exceptions.
    \item \textbf{$\mu$-Initiality}: $\mu$ is \textit{the} unique instruction-consistent cost measure, not just one of many.
    \item \textbf{$\mu$-Hierarchy}: For every $k \ge 1$, level-$k$ certification costs exactly $k$. No shortcut closes the gap.
    \item \textbf{$\mu$-Ledger Necessity}: The ledger is provably inaccessible to classical observers. No function of (mem, regs, pc) can recover $\mu$, certification, or the categorical structure (vm\_graph). The separation holds from every reachable state---\path{coq/NecessityOfMuLedger.v} and \path{coq/kernel/NecessityAbstract.v}.
    \item \textbf{Categorical Separation}: Two states with identical classical shadow can differ in morphism structure. No classical probe sees the difference---\path{coq/kernel/PartitionSeparation.v}.
    \item \textbf{Structural Entitlement Representation}: Strict feasible-set narrowing plus a distinguishing observation yields a strictly stronger predicate, a structure-addition event, and a $\Delta\mu$ lower bound.
    \item \textbf{Classical CHSH Bound and Tsirelson bound}: $|S| \le 2$ for local strategies; $|S| \le 2\sqrt{2}$ for the zero-marginal NPA model. Both are pure algebraic results, no physics.
    \item \textbf{Thiele honesty $\iff$ NPA realizability}: A zero-axiom biconditional. The machine's column-contractivity conditions are exactly the zero-marginal NPA polynomial realizability conditions---\path{coq/kernel/QuantumPartitionPSD.v}, corollary \texttt{column\_contractive\_iff\_quantum\_realizable}.
    \item \textbf{Discrete angle-defect identity (planar triangulations)}: Under the predicate \texttt{well\_formed\_triangulated} (defined in \path{coq/kernel/DiscreteTopology.v} and used by \path{coq/kernel/DiscreteGaussBonnet.v}), the angle defect of an equilateral triangulation equals $5\pi \cdot \chi(g)$. This is \emph{not} classical Gauss-Bonnet (which is $2\pi\chi$ for closed surfaces); the predicate restricts to a planar-disk family.
    \item \textbf{Turing Strictness}: The Thiele Machine strictly extends classical computation: states reachable by Thiele instructions cannot be reached by any classical program of any length---\path{coq/kernel/TuringStrictness.v}.
\end{itemize}

\textbf{What is built}:
\begin{itemize}
    \item Coq formal kernel (active proof tree, zero admits, zero global \texttt{Axiom}/\texttt{Parameter} declarations in the active audited tree; the two named Section-Variable trust boundaries are listed in the ``What is proven'' parenthetical above)
    \item Python reference VM with two-pass assembler (\path{scripts/thiele_asm.py})
    \item Synthesizable Verilog RTL extracted from Coq through Kami
    \item Automated tests enforcing opcode completeness, certificate plumbing, and cross-layer comparison gates
    \item 46 opcodes, all in hardware, all with formal correctness proofs: 36 unconditional and 10 carrying Qed bisimulation proofs under inductive structural invariants --- 3 (PNEW, PSPLIT, PMERGE) under partition-table well-formedness plus arithmetic preconditions, and 7 (MORPH, MORPH\_ID, MORPH\_DELETE, MORPH\_ASSERT, MORPH\_GET, COMPOSE, MORPH\_TENSOR) under the five-conjunct \texttt{extended\_hw\_invariant} ($\texttt{pt\_well\_formed} \wedge \texttt{morph\_table\_wf} \wedge \texttt{coupling\_zero\_empty} \wedge \texttt{coupling\_wf} \wedge \texttt{coupling\_desc\_safe}$), proven preserved by every \texttt{kami\_step}. Zero structural gaps (\path{coq/kami_hw/RTLGapRegistry.v}, theorem \texttt{rtl\_coverage\_partition}).
\end{itemize}

\textbf{What this project is not}: It is not a claim that I solved P vs NP. It is not a theory of everything. It is not a claim that computation is physics or that physics is computation. The CHSH algebraic result is pure math. The Landauer connection is motivation. That is the full extent of the physics content here.

If you can find an error, find it. Everything is open source, documented, and testable.

\vspace{1cm}
\noindent\textbf{Keywords:} Formal Verification, Coq, Computational Complexity, Information Theory, Hardware Synthesis, Partition Logic

% Table of Contents
\onecolumn
\tableofcontents
\twocolumn

\chapter{Introduction}
\input{chapters/01_introduction}

\chapter{Background and Related Work}
\input{chapters/02_background}

\chapter{Theory: The Thiele Machine Model}
\input{chapters/03_theory}

\chapter{Implementation: The 3-Layer Comparison}
\input{chapters/04_implementation}

\chapter{Verification: The Coq Proofs}
\input{chapters/05_verification}

\chapter{Evaluation: Empirical Evidence}
\input{chapters/06_evaluation}

\chapter{Discussion: Implications and Future Work}
\input{chapters/07_discussion}

\chapter{Conclusion}
\input{chapters/08_conclusion}

\appendix

\chapter{The Verifier System}
\input{chapters/09_verifier_system}

\chapter{Experimental Validation Suite}
\input{chapters/11_experiments}

\chapter{Hardware Implementation and Demonstrations}
\input{chapters/13_hardware_and_demos}

\chapter{Glossary of Terms}
\input{appendix/glossary}

\bibliographystyle{plain}
\bibliography{references}

\end{document}
